Maximum Entropy Spectral Analysis: an application to gravitational waves data analysis

### 2024-07-01 paper/mesa.tex
Modified paper for EPJ

\documentclass{aa}
\bibpunct{(}{)}{;}{a}{}{,}
\usepackage[varg]{txfonts}
\usepackage{amsmath,amssymb}
\usepackage{amsfonts}
\usepackage[normalem]{ulem}
\usepackage{textcomp}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{bm}
\usepackage{afterpage}
\usepackage{graphicx}
\usepackage{psfrag}
\usepackage{mathtools}
%\usepackage{tensor}
\usepackage{layouts}
%\usepackage{DejaVuSans}
\usepackage{epstopdf}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{ragged2e}
\usepackage{cleveref}
\usepackage{blindtext}
\usepackage{graphicx}
\usepackage{siunitx}
	\sisetup{output-decimal-marker={.}}
\usepackage{float}

	%some math symbols
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\DeclareMathOperator{\sign}{sign}
\renewcommand{\d}[1]{\ensuremath{\operatorname{d}\!{#1}}}
%argmin and argmax
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

% comments command
\newcommand{\wdp}[1]{{\textcolor{magenta}{\texttt{walter: #1}} }}
\newcommand{\amartini}[1]{{\textcolor{blue}{\texttt{alessandro: #1}} }}
\newcommand{\sschmidt}[1]{{\textcolor{purple}{\texttt{stefano: #1}} }}
\newcommand{\todo}[1]{{\textcolor{red}{\texttt{TODO: #1}} }}

@@

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TITLE
	\title{Maximum Entropy Spectral Analysis: an application to gravitational waves data analysis}
	\author{Alessandro~{Martini}\inst{1,2,3}\thanks{alessandro.martini-1@unitn.it} \and 
	Stefano~{Schmidt}\inst{4,5}\thanks{s.schmidt@uu.nl} \and Gregory~{Ashton}\inst{6} \and Walter~{Del Pozzo}\inst{1}\thanks{walter.delpozzo@unipi.it}}

@@

  \abstract{
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR($p$) of order $p$.
A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
We provide a ready-to-use implementation of the algorithm in the form of a python package \texttt{memspectrum} and we characterize our implementation by performing a power spectral density analysis on synthetic data (with known power spectral density) and by comparing different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the ubiquitous Welch algorithm, using synthetic data generated from the GW150914 strain spectrum released  by the LIGO-Virgo-Kagra collaboration.
We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
systematically lower variance and bias. This is particularly manifest in the case of a small ($O(5000)$) number of data points, making Burg's method most suitable to work in this regime. Since this is close to the typical length of analysed gravitational waves data, improving the estimate of the PSD in this regime leads to more reliable posterior profiles for the system under study.
We conclude our investigation by utilising MESA, and its particularly easy parametrisation where the only free parameter is the order $p$ of the AR process, to marginalise over the interferometers noise PSD in conjunction with inferring the parameters of GW150914.}

\titlerunning{MESA: a case study}
\authorrunning{Martini, Schmidt, Del Pozzo}

@@

The problem of inferring the morphology and the defining parameters of deterministic signals superimposed to stochastic processes is one of the most wide spread and interesting problems in several areas of human activities. Whenever some form of model for the signal we are looking for is available, the problem is typically solved via the Wiener filter, defined as the whitening filter that maximises the signal-to-noise ratio, i.e. the relative power of the (known) signal over the power of the (known) underlying stochastic process. Hence, signal detection and characterisation requires accurate knowledge of i) the shape of the signal we are looking for and ii) the statistical properties of the stochastic process. The construction of signal models is typically driven either by physical or by mathematical arguments hence, although extremely difficult in general, it is doable. On the other hand, stochastic process models can be extremely difficult to construct, both for practical and theoretical reasons. A stochastic process is fully described by the knowledge of the probability distribution governing its realisations -- the ``paths'' of the random variable under scrutiny -- over the entire time axis, from $t=-\infty$ to $t=\infty$. Clearly this is not possible in practice. Therefore modelling a stochastic process either relies on modelling of the underlying physical processes, thus falling back onto the deterministic case, or on modelling the mathematical and statistical properties of the process, and potentially infering them from the process realisations. The study of the properties of stochastic processes is thus a crucial task in many fields of physics, astronomy, quantitative biology, as well as engineering and finance. Among the classes of stochastic processes, a key role is played by \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to translation of the independent variable, usually the time $t$. If $x(t)$ is a wide-sense stationary process, its statistical properties are completely determined by the knowledge of the (many-points) autocorrelation functions. In practice, one often has easy access to the two-point correlation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or, equivalently, to the process \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem, in wide-sense stationary processes, the two are in fact related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In the context of gravitational waves physics, e.g. ~\citep{Finn_1992}, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will be explored further in what follows. The latter definition in Eq. ~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.

An ubiquitous method for such a computation is due to \citet{Welch1967} and it is based on Eqs. (\ref{eq:psd-autocorrelation}-\ref{eq:psd-f-definition}).
The PSD is obtained by slicing the observed realisation $x(t_1),\ldots,x(t_n)$ of the process $x(t)$ into many window-corrected batches and averaging the squared moduli of their Fourier transforms.
This approach is equivalent~\citep{Lomb,Scargle} to taking the Fourier Transform of the windowed sample autocorrelation $\rho_W$, written as
\begin{equation}
    \rho_{W} = \left\{W_0\rho_0,W_{\pm 1}\rho_{\pm 1}, \dots, W_{\pm M}\rho_{\pm M}, 0, 0, \dots \right\},
\end{equation}
where $\rho$ is the empirical autocorrelation and $M$ is the maximum time lag at which the autocorrelation is computed.
The sequence $W$ is a window function that can be chosen in several different ways, each choice presenting advantages and disadvantages for the final estimate of the PSD.

@@

A alternative approach providing a smooth PSD estimation, is to adopt a parametric model for the PSD and to fit its parameters to the data with a Reversible Jump Markov Chain Monte Carlo \citep[e.g.]{Cornish_2015, Littenberg_2015}. Despite being effective, this method is problem dependent, since it needs to make definite assumptions on the shape of the PSD. Moreover, it can be computationally expensive and it does not come with a handy implementation available to the public. For all the above reasons, we did not consider such methods in our work.

An appealing alternative, based on the Maximum Entropy principle \citep{JaynesArticle,jaynes2003ptl, Jaynes_MAXENT}, has been derived by~\citet{burg1975maximum}. Being rooted on solid theoretical foundations, we will see that Burg's method, unlike Welch's, does not require any preprocessing of the data and requires very little tuning of the algorithm parameters, since it provides an iterative closed form expression for the spectrum of a stochastic stationary time series. Furthermore, it embeds the PSD estimation problem into an elegant theoretical framework and makes minimal assumptions on the nature of the data.
Lastly and most importantly, it provides a robust link between spectral density estimation and the field of autoregressive processes. This provides a natural and simple machinery to forecast a time series, thus predicting future observations based on previous ones.

@@

\section{Theoretical foundations}\label{sec:foundations}
The Maximum Entropy principle (MAXENT) is among the most important results in probability theory. It provides a way to uniquely assign probabilities to a phenomenon in a way that best represent our state of knowledge, while being non-committal to unavailable information. Its domain of application turned out to be wider than expected. In fact, thanks to~\citet{burg1975maximum}, this method has also been applied to perform high quality computation of power spectral densities of time series.

@@

Before introducing the MAXENT principle, we will define, via some simple examples, the two core concepts of the problem and the roles they play in deductive inference: the `evidence' and the `information'.
Let us start with the `information' (or information entropy): it is a measure of the degree of uncertainty on the outcomes of some experiment and specifies the length of the message necessary to provide a full description of the system under study. As an example, no information is brought if we are studying a system whose outcome is certain (the outcome is known with probability $p = 1$), as in this case, a communication is not even needed.
\citet{Shannon} proposed the quantity
\begin{equation}\label{eq:information}
    I = \log_2 \frac{1}{p(x)}
\end{equation}
to measure the quantity of information brought by an outcome $x$ with probability $p(x)$. It is additive quantity as well as a monotonically decreasing function of $p \in [0, 1]$: the more uncertain the outcome, the higher the information it brings.

@@

\citet{Shannon} showed that the only functional form satisfying continuity with respect to its parameters, additivity and that has a maximum for equal probability events is:
\begin{equation}\label{eq:entropy}
    H[p_1, \dots, p_N] = - \sum_{i = 1}^N p_i\log{p_i}\,,
\end{equation}
which can be interpreted as the `expected information' brought by an experiment with $N$ possible outcomes each with its own probability $p_i$.
In the continuous case:
\begin{equation} \label{eq:entropy_continuos}
    H[p(x)] = - \int p(x)\ln p(x) dx,
\end{equation}
We call the functional $H$ information entropy\footnote{In defining the information entropy as in Eq.~\eqref{eq:entropy_continuos}, we are implicitly assuming a uniform measure over the parameter space. In case of a non-uniform measure $m(x)$, the definition generalises to $ H[p(x)] = - \int p(x)\ln \frac{p(x)}{m(x)} dx$.}.

We now turn to the core of our problem: how can we assign probabilities to a set of events keeping into account our knowledge of the system and, at the same time, ensure it is non-committal towards unavailable knowledge?
The ``knowledge'' at our disposal about the system under investigation is what we define `evidence' and any probability assignment is given such evidence, in agreement with \citet{Cox} construction of probability. In the case above, our knowledge on the system is only the total number $N$ of different outcomes -- this is a minimal requirement. Of course, more complex evidence constraints can be applied.

It is very common that the constraints provided by the evidence are not enough for setting the probabilities for each event: in this case, it is reasonable to assume that the probability assignment should make the experiment as unpredictable as possible\footnote{
In~\citet{Jaynes_MAXENT} this statement is made more precise and justified more thoroughly, with arguments based on combinatorial analysis.
}.
In other words, the information entropy content introduced by the probability assignment should be as large as possible, in accordance with the available evidence. MAXENT formalises this reasoning by stating that probabilities should be assigned by maximizing uncertainty (information entropy) using evidence as a constraint. 
This defines a variational problem, where the information entropy functional $H\left[p_1, \dots, p_N\right]$, defined in Eq. ~\eqref{eq:entropy}, has to be maximized.

@@

For this work, we are especially interested in the multi-dimensional case. Suppose we have a vector of measurements $(x(t_1),\ldots,x(t_n)) = (x_1, \ldots, x_n)$ that we conveniently express as a single realization of an unknown stochastic process $x(t)$ and we have information about the expectation value of the process $\mu(t)$ and on the matrix of autocovariances $C_{ij} \equiv C(t_i,t_j)$, then the MAXENT distribution is the $n$-dimensional multivariate Gaussian distribution~\citep{gregory_2005}: 
\begin{align}
    p&\left((x_1, \ldots, x_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det C\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{i,j}(x_i-\mu_i) (x_j-\mu_j)C^{-1}_{ij} \right)\,. 
\end{align}

For a wide-sense stationary process the mean function is independent of time, hence it can be redefined to be equal to zero without loss of generality, and the auto-covariance function is dependent only on the time lag $\tau \equiv t_i - t_j$. One can thus choose a sampling rate $\Delta t$ so that $C_{ij} = C((i-j)\Delta t)$. The autocovariance matrix thus becomes a Toeplitz matrix\footnote{
We remind the reader that a Toeplitz matrix is a matrix in the form:
$\begin{pmatrix}
	a_0 & a_1 & a_2 & \ldots & \ldots& \ldots &a_n\\
	a_{-1} & a_0 & a_1 & \ldots & \ldots& \ldots &a_{n-1}\\
	a_{-2} & a_{-1} & a_0 & \ldots & \ldots & \ldots  &a_{n-2}\\
	\vdots & \vdots & \vdots & \vdots & \vdots & \vdots &\vdots\\
	a_{-n +1} & \ldots & \ldots & \ldots& a_{-1} & a_0    &a_{1}\\
	a_{-n} & \ldots & \ldots & \ldots& a_{-2} & a_{-1} & a_0
\end{pmatrix}$
}.
Toeplitz matrices are asymptotically equivalent to circulant matrices and thus diagonalized by the discrete Fourier transform base~\citep{Gray}.
%R. M. Gray, Foundations and Trends⃝R in Communications and Information Theory 2, 155 (2005), URL https://doi.org/10.1561/0100000006.
Some simple algebra shows that the time-domain multivariate Gaussian can be transformed into the equivalent frequency domain 
probability distribution:
\begin{align}\label{eq:Whittle}
p&\left((\tilde{x}_1, \ldots, \tilde{x}_{n/2})\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det S\right)^{n / 2}}\exp\left(-\frac{1}{2}\sum_{ij}\tilde{x}_i S^{-1}_{ij} \tilde{x}_j \right)\,,
\end{align}
where the matrix $S_{ij} = S_i \delta_{ij}$ is an $n\times n$ diagonal matrix whose elements are the PSD $S(f)$ calculated at frequency $f_i$.
Many readers will recognise the familiar form of the Whittle likelihood that stands at the basis of the \emph{matched filter} method \citep{prob_information_theory}
and of gravitational waves data analysis, \citep[e.g.]{Finn_1992, FINDCHIRP}.
Thanks to MAXENT, the problem of defining the probability distribution describing a wide-sense stationary process is thus 
entirely reduced to the estimation of the PSD or, equivalently, the autocovariance function.

@@

Before doing so, there is a technicality to solve: the definition of entropy depends on a probability distribution, not on the PSD.
It can be shown ~\citep[e.g.]{AblesMESA, Bartlett} that the variational problem can be formulated in terms of the power spectral density $S(f)$ alone by considering our signal as the result of the filtering a white noise process using a filter with transfer function $T(f)$ equal to $S(f)$\footnote{
A filter with transfer function $T(f)$ takes in input a time series $x_t$ and outputs a times series $y_t$ such that:
$$T(f) = \frac{\tilde{y}(f)}{\tilde{x}(f)}$$
where $\tilde{x}(f)$ denotes the Fourier transform of $x_t$ (and similarly for $y_t$)
}.
The difference in entropy between the input and the output time series (i.e. the entropy gain) obtained by such filter applied on white noise is:
\begin{equation}\label{eq:EntropyGain}
    \Delta H = \int_{- Ny}^{Ny}\log S(f) df\,.
\end{equation}
where $\Delta t$ is sampling rate and $Ny \equiv \frac{1}{2 \Delta t}$  is the Nyquist frequency.
Thus maximising Eq.~\eqref{eq:EntropyGain} is equivalent to maximizing Eq.~\eqref{eq:entropy_continuos}.

@@

The application of MESA is not limited to spectral estimates, but it also provides a link between spectral analysis and the study
of autoregressive processes (AR)~\citep{doi:10.1029/RG013i001p00183}.
An autoregressive stationary process of order $p$, AR($p$), is a time series whose values satisfy the following expression: 
\begin{equation} \label{eq:AR_p}
    x_t - b_1 x_{t-1} - b_2 x_{t-2} \dots b_p x_{t - p} = \nu_t
\end{equation}
where $b_1, \ldots, b_p$ are real coefficients and $\nu_t$ is white noise with a given variance $\sigma^2$.
Thus, an AR($p$) process models the dependence of the value of the process at time $t$ from the last $p$ observations, 
thus being potentially able to model complex autocorrelation structures within observations.

>>>>

\documentclass[epj,nopacs]{svjour}
\usepackage[varg]{txfonts}
\let\iint\relax
\let\iiint\relax
\let\iiiint\relax
\let\idotsint\relax
\usepackage{amsmath,amssymb}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage[normalem]{ulem}
\usepackage{textcomp}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{bm}
\usepackage{afterpage}
\usepackage{graphicx}
\usepackage{psfrag}
\usepackage{mathtools}
%\usepackage{tensor}
\usepackage{layouts}
%\usepackage{DejaVuSans}
\usepackage{epstopdf}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{ragged2e}
%\usepackage{cleveref}
\usepackage{blindtext}
\usepackage{graphicx}
\usepackage{siunitx}
	\sisetup{output-decimal-marker={.}}
\usepackage{float}
%\usepackage{natbib}

	%some math symbols
%\newcommand{\R}{\mathbb{R}}
%\newcommand{\N}{\mathbb{N}}
\DeclareMathOperator{\sign}{sign}
%\renewcommand{\d}[1]{\ensuremath{\operatorname{d}\!{#1}}}
%argmin and argmax
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\usepackage{cite}

% comments command
%\newcommand{\wdp}[1]{{\textcolor{magenta}{\texttt{walter: #1}} }}
%\newcommand{\amartini}[1]{{\textcolor{blue}{\texttt{alessandro: #1}} }}
%\newcommand{\sschmidt}[1]{{\textcolor{purple}{\texttt{stefano: #1}} }}
%\newcommand{\todo}[1]{{\textcolor{red}{\texttt{TODO: #1}} }}

@@

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TITLE
	\title{Maximum Entropy Spectral Analysis: an application to gravitational waves data analysis}
	\author{Alessandro~{Martini}\inst{1,2,3}\thanks{alessandro.martini-1@unitn.it} \and 
	Stefano~{Schmidt}\inst{4,5}\thanks{s.schmidt@uu.nl} \and Gregory~{Ashton}\inst{6} \thanks{Gregory.Ashton@rhul.ac.uk}\and Walter~{Del Pozzo}\inst{1}\thanks{walter.delpozzo@unipi.it}}

@@

\abstract{
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, offers a powerful tool for spectral estimation of a time-series. It relies on Jaynes’ maximum entropy principle, allowing the spectrum of a stochastic process to be inferred using the coefficients of an autoregressive process AR($p$) of order $p$. A closed-form recursive solution provides estimates for both the autoregressive coefficients and the order $p$ of the process. We provide a ready-to-use implementation of this algorithm in a Python package called \texttt{memspectrum}, characterized through power spectral density (PSD) analysis on synthetic data with known PSD and comparisons of different criteria for stopping the recursion. Additionally, we compare the performance of our implementation with the ubiquitous Welch algorithm, using synthetic data generated from the GW150914 strain spectrum released by the LIGO-Virgo-Kagra collaboration. Our findings indicate that Burg’s method provides PSD estimates with systematically lower variance and bias. This is particularly manifest in the case of a small (O($5000$)) number of data points, making Burg’s method most suitable to work in this regime. Since this is close to the typical length of analysed gravitational waves data, improving
the estimate of the PSD in this regime leads to more reliable posterior profiles for the system under study. We conclude
our investigation by utilising MESA, and its particularly easy parametrisation where the only free parameter is the order
p of the AR process, to marginalise over the interferometers noise PSD in conjunction with inferring the parameters of
GW150914 
}

%  \abstract{
%The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR($p$) of order $p$.
%A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
%We provide a ready-to-use implementation of the algorithm in the form of a python package \texttt{memspectrum} and we characterize our implementation by performing a power spectral density analysis on synthetic data (with known power spectral density) and by comparing different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the ubiquitous Welch algorithm, using synthetic data generated from the GW150914 strain spectrum released  by the LIGO-Virgo-Kagra collaboration.
%We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
%systematically lower variance and bias. This is particularly manifest in the case of a small ($O(5000)$) number of data points, making Burg's method most suitable to work in this regime. Since this is close to the typical length of analysed gravitational waves data, improving the estimate of the PSD in this regime leads to more reliable posterior profiles for the system under study.
%We conclude our investigation by utilising MESA, and its particularly easy parametrisation where the only free parameter is the order $p$ of the AR process, to marginalise over the interferometers noise PSD in conjunction with inferring the parameters of GW150914.}

\titlerunning{MESA: an application to gravitational waves data analysis}
\authorrunning{Martini, Schmidt, Ashton, Del Pozzo}

@@

The problem of inferring the morphology and the defining parameters of deterministic signals superimposed to stochastic processes is one of the most wide spread and interesting problems in several areas of human activities. Whenever some form of model for the signal we are looking for is available, the problem is typically solved via the Wiener filter, defined as the whitening filter that maximises the signal-to-noise ratio, i.e. the relative power of the (known) signal over the power of the (known) underlying stochastic process. Hence, signal detection and characterisation requires accurate knowledge of i) the shape of the signal we are looking for and ii) the statistical properties of the stochastic process. The construction of signal models is typically driven either by physical or by mathematical arguments hence, although extremely difficult in general, it is doable. On the other hand, stochastic process models can be extremely difficult to construct, both for practical and theoretical reasons. A stochastic process is fully described by the knowledge of the probability distribution governing its realisations -- the ``paths'' of the random variable under scrutiny -- over the entire time axis, from $t=-\infty$ to $t=\infty$. Clearly this is not possible in practice. Therefore modelling a stochastic process either relies on modelling of the underlying physical processes, thus falling back onto the deterministic case, or on modelling the mathematical and statistical properties of the process, and potentially infering them from the process realisations. The study of the properties of stochastic processes is thus a crucial task in many fields of physics, astronomy, quantitative biology, as well as engineering and finance. Among the classes of stochastic processes, a key role is played by \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to translation of the independent variable, usually the time $t$. If $x(t)$ is a wide-sense stationary process, its statistical properties are completely determined by the knowledge of the (many-points) autocorrelation functions. In practice, one often has easy access to the two-point correlation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or, equivalently, to the process \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem, in wide-sense stationary processes, the two are in fact related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In the context of gravitational waves physics, e.g. \cite{Finn_1992}, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will be explored further in what follows. The latter definition in Eq. ~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.

An ubiquitous method for such a computation is due to \cite{Welch1967} and it is based on Eqs. (\ref{eq:psd-autocorrelation}-\ref{eq:psd-f-definition}).
The PSD is obtained by slicing the observed realisation $x(t_1),\ldots,x(t_n)$ of the process $x(t)$ into many window-corrected batches and averaging the squared moduli of their Fourier transforms.
This approach is equivalent~\cite{Lomb}\cite{Scargle} to taking the Fourier Transform of the windowed sample autocorrelation $\rho_W$, written as
\begin{equation}
    \rho_{W} = \left\{W_0\rho_0,W_{\pm 1}\rho_{\pm 1}, \dots, W_{\pm M}\rho_{\pm M}, 0, 0, \dots \right\},
\end{equation}
where $\rho$ is the empirical autocorrelation and $M$ is the maximum time lag at which the autocorrelation is computed.
The sequence $W$ is a window function that can be chosen in several different ways, each choice presenting advantages and disadvantages for the final estimate of the PSD.

@@

A alternative approach providing a smooth PSD estimation, is to adopt a parametric model for the PSD and to fit its parameters to the data with a Reversible Jump Markov Chain Monte Carlo \cite{Cornish_2015,Littenberg_2015}. Despite being effective, this method is problem dependent, since it needs to make definite assumptions on the shape of the PSD. Moreover, it can be computationally expensive and it does not come with a handy implementation available to the public. For all the above reasons, we did not consider such methods in our work.

An appealing alternative, based on the Maximum Entropy principle \cite{JaynesArticle, jaynes2003ptl, Jaynes_MAXENT}, has been derived by~\cite{burg1975maximum}. Being rooted on solid theoretical foundations, we will see that Burg's method, unlike Welch's, does not require any preprocessing of the data and requires very little tuning of the algorithm parameters, since it provides an iterative closed form expression for the spectrum of a stochastic stationary time series. Furthermore, it embeds the PSD estimation problem into an elegant theoretical framework and makes minimal assumptions on the nature of the data.
Lastly and most importantly, it provides a robust link between spectral density estimation and the field of autoregressive processes. This provides a natural and simple machinery to forecast a time series, thus predicting future observations based on previous ones.

@@

\section{Theoretical foundations}\label{sec:foundations}
The Maximum Entropy principle (MAXENT) is among the most important results in probability theory. It provides a way to uniquely assign probabilities to a phenomenon in a way that best represent our state of knowledge, while being non-committal to unavailable information. Its domain of application turned out to be wider than expected. In fact, thanks to~\cite{burg1975maximum}, this method has also been applied to perform high quality computation of power spectral densities of time series.

@@

Before introducing the MAXENT principle, we will define, via some simple examples, the two core concepts of the problem and the roles they play in deductive inference: the `evidence' and the `information'.
Let us start with the `information' (or information entropy): it is a measure of the degree of uncertainty on the outcomes of some experiment and specifies the length of the message necessary to provide a full description of the system under study. As an example, no information is brought if we are studying a system whose outcome is certain (the outcome is known with probability $p = 1$), as in this case, a communication is not even needed.
\cite{Shannon} proposed the quantity
\begin{equation}\label{eq:information}
    I = \log_2 \frac{1}{p(x)}
\end{equation}
to measure the quantity of information brought by an outcome $x$ with probability $p(x)$. It is additive quantity as well as a monotonically decreasing function of $p \in [0, 1]$: the more uncertain the outcome, the higher the information it brings.

@@

\cite{Shannon} showed that the only functional form satisfying continuity with respect to its parameters, additivity and that has a maximum for equal probability events is:
\begin{equation}\label{eq:entropy}
    H[p_1, \dots, p_N] = - \sum_{i = 1}^N p_i\log{p_i}\,,
\end{equation}
which can be interpreted as the `expected information' brought by an experiment with $N$ possible outcomes each with its own probability $p_i$.
In the continuous case:
\begin{equation} \label{eq:entropy_continuos}
    H[p(x)] = - \int p(x)\ln p(x) dx,
\end{equation}
We call the functional $H$ information entropy\footnote{In defining the information entropy as in Eq.~\eqref{eq:entropy_continuos}, we are implicitly assuming a uniform measure over the parameter space. In case of a non-uniform measure $m(x)$, the definition generalises to $ H[p(x)] = - \int p(x)\ln \frac{p(x)}{m(x)} dx$.}.

We now turn to the core of our problem: how can we assign probabilities to a set of events keeping into account our knowledge of the system and, at the same time, ensure it is non-committal towards unavailable knowledge?
The ``knowledge'' at our disposal about the system under investigation is what we define `evidence' and any probability assignment is given such evidence, in agreement with \cite{Cox} construction of probability. In the case above, our knowledge on the system is only the total number $N$ of different outcomes -- this is a minimal requirement. Of course, more complex evidence constraints can be applied.

It is very common that the constraints provided by the evidence are not enough for setting the probabilities for each event: in this case, it is reasonable to assume that the probability assignment should make the experiment as unpredictable as possible\footnote{
In~\cite{Jaynes_MAXENT} this statement is made more precise and justified more thoroughly, with arguments based on combinatorial analysis.
}.
In other words, the information entropy content introduced by the probability assignment should be as large as possible, in accordance with the available evidence. MAXENT formalises this reasoning by stating that probabilities should be assigned by maximizing uncertainty (information entropy) using evidence as a constraint. 
This defines a variational problem, where the information entropy functional $H\left[p_1, \dots, p_N\right]$, defined in Eq. ~\eqref{eq:entropy}, has to be maximized.

@@

For this work, we are especially interested in the multi-dimensional case. Suppose we have a vector of measurements $(x(t_1),\ldots,x(t_n)) = (x_1, \ldots, x_n)$ that we conveniently express as a single realization of an unknown stochastic process $x(t)$ and we have information about the expectation value of the process $\mu(t)$ and on the matrix of autocovariances $C_{ij} \equiv C(t_i,t_j)$, then the MAXENT distribution is the $n$-dimensional multivariate Gaussian distribution~\cite{gregory_2005}: 
\begin{align}
    p&\left((x_1, \ldots, x_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det C\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{i,j}(x_i-\mu_i) (x_j-\mu_j)C^{-1}_{ij} \right)\,. 
\end{align}

For a wide-sense stationary process the mean function is independent of time, hence it can be redefined to be equal to zero without loss of generality, and the auto-covariance function is dependent only on the time lag $\tau \equiv t_i - t_j$. One can thus choose a sampling rate $\Delta t$ so that $C_{ij} = C((i-j)\Delta t)$. The autocovariance matrix thus becomes a Toeplitz matrix\footnote{
We remind the reader that a Toeplitz matrix is a matrix in the form:
$\begin{pmatrix}
	a_0 & a_1 & a_2 & \ldots & \ldots& \ldots &a_n\\
	a_{-1} & a_0 & a_1 & \ldots & \ldots& \ldots &a_{n-1}\\
	a_{-2} & a_{-1} & a_0 & \ldots & \ldots & \ldots  &a_{n-2}\\
	\vdots & \vdots & \vdots & \vdots & \vdots & \vdots &\vdots\\
	a_{-n +1} & \ldots & \ldots & \ldots& a_{-1} & a_0    &a_{1}\\
	a_{-n} & \ldots & \ldots & \ldots& a_{-2} & a_{-1} & a_0
\end{pmatrix}$
}.
Toeplitz matrices are asymptotically equivalent to circulant matrices and thus diagonalized by the discrete Fourier transform base~\cite{Gray}.
%R. M. Gray, Foundations and Trends⃝R in Communications and Information Theory 2, 155 (2005), URL https://doi.org/10.1561/0100000006.
Some simple algebra shows that the time-domain multivariate Gaussian can be transformed into the equivalent frequency domain 
probability distribution:
\begin{align}\label{eq:Whittle}
p&\left((\tilde{x}_1, \ldots, \tilde{x}_{n/2})\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det S\right)^{n / 2}}\exp\left(-\frac{1}{2}\sum_{ij}\tilde{x}_i S^{-1}_{ij} \tilde{x}_j \right)\,,
\end{align}
where the matrix $S_{ij} = S_i \delta_{ij}$ is an $n\times n$ diagonal matrix whose elements are the PSD $S(f)$ calculated at frequency $f_i$.
Many readers will recognise the familiar form of the Whittle likelihood that stands at the basis of the \emph{matched filter} method \cite{prob_information_theory}
and of gravitational waves data analysis, \cite[e.g.]{Finn_1992,FINDCHIRP}.
Thanks to MAXENT, the problem of defining the probability distribution describing a wide-sense stationary process is thus 
entirely reduced to the estimation of the PSD or, equivalently, the autocovariance function.

@@

Before doing so, there is a technicality to solve: the definition of entropy depends on a probability distribution, not on the PSD.
It can be shown ~\cite[e.g.]{AblesMESA,Bartlett} that the variational problem can be formulated in terms of the power spectral density $S(f)$ alone by considering our signal as the result of the filtering a white noise process using a filter with transfer function $T(f)$ equal to $S(f)$\footnote{
A filter with transfer function $T(f)$ takes in input a time series $x_t$ and outputs a times series $y_t$ such that:
$$T(f) = \frac{\tilde{y}(f)}{\tilde{x}(f)}$$
where $\tilde{x}(f)$ denotes the Fourier transform of $x_t$ (and similarly for $y_t$)
}.
The difference in entropy between the input and the output time series (i.e. the entropy gain) obtained by such filter applied on white noise is:
\begin{equation}\label{eq:EntropyGain}
    \Delta H = \int_{- Ny}^{Ny}\log S(f) df\,.
\end{equation}
where $\Delta t$ is sampling rate and $Ny \equiv \frac{1}{2 \Delta t}$  is the Nyquist frequency.
Thus maximising Eq.~\eqref{eq:EntropyGain} is equivalent to maximizing Eq.~\eqref{eq:entropy_continuos}.

@@

The application of MESA is not limited to spectral estimates, but it also provides a link between spectral analysis and the study
of autoregressive processes (AR)~\cite{doi:10.1029/RG013i001p00183}.
An autoregressive stationary process of order $p$, AR($p$), is a time series whose values satisfy the following expression: 
\begin{equation} \label{eq:AR_p}
    x_t - b_1 x_{t-1} - b_2 x_{t-2} \dots b_p x_{t - p} = \nu_t
\end{equation}
where $b_1, \ldots, b_p$ are real coefficients and $\nu_t$ is white noise with a given variance $\sigma^2$.
Thus, an AR($p$) process models the dependence of the value of the process at time $t$ from the last $p$ observations, 
thus being potentially able to model complex autocorrelation structures within observations.
